Normalization by Evaluation for \\ Intuitionistic Propositional Logic

### 2018-07-19 notes/nbeSum.tex
ITC: bibliography
ITC: spell-check

\nonstopmode
\documentclass[a4paper]{article}
\usepackage{natbib}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
%\usepackage{dsfont}

@@

@@

Indices $x$ into the context $\Gamma$ of hypothesis, locating hypothesis
$A$, are written $x : \Hyp[\Gamma]A$ and defined inductively by the
following constructors:
\[
  \ru{}{\ttop : \Hyp[\Gamma.A]A}
\qquad
  \ru{x : \Hyp[\Gamma]A
    }{\pop[B] x : \Hyp[\Gamma.B]A}
\]
Deriviations $t$ of the truth of a formula $A$ under assumptions
$\Gamma$ (judgement $\Gamma \der A$)
are written $t : \Tm[\Gamma]A$ and defined inductively as follows:

@@

The following indexed grammar for \emph{normal} derivations $\NfG A$ of
judgement $\Gamma \der A$ ensures the subformula property.  It is
defined mutually with a grammar $\NeG A$ of \emph{neutral} derivations.
Neutral derivations are ``straightline'' consequences of hypotheses
(without case distinction or absurdity elimination).

@@

\begin{exercise}
  Using the constructions in this section, show that the category of
  presheaves and natural transformations is cartesian closed.
\end{exercise}

@@

Instead of defining the semantics $\den A$ of formula $A$ by induction
on $A$, we can consider a category whose objects are interpretations
of formulas, and show that this category is cartesian closed.

An object in this category would be a quadruple consisting of:
\[
\begin{array}{lll@{\qquad}l}
  \A & : & \PSh  & \mbox{a presheaf} \\
  A  & : & \Form & \mbox{a formula} \\
  \treflect^A & : & \Ne A \todot \A
    & \mbox{a method to reflect neutrals into $\A$} \\
  \treify^A & : & \A \todot \Nf A
    & \mbox{a method to reify values of $\A$} \\
\end{array}
\]
\begin{exercise}
  Complete this definition by a suitable notion of morphism.
\end{exercise}
\begin{exercise}
  Show that this category is cartesian closed.
\end{exercise}

@@

Given a presheaf $\A$, we define a new presheaf $\Cover \A$
inductively by the following constructors:
\begin{gather*}
  \ru{a : \A_\Gamma}{\returnC a : \Cover\A_\Gamma}
\qquad
  \ru{t : \NeG \bot}{\falseC t : \Cover\A_\Gamma}
\\[1ex]
  \ru{t : \NeG{C \lor D} \qquad
      c : \Cover\A_{\Gamma.C} \qquad
      d : \Cover\A_{\Gamma.D}
    }{\orC t c d : \Cover\A_\Gamma}
\end{gather*}
\begin{exercise}
  Show that $\Cover\A$ is indeed a presheaf.
\end{exercise}
We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
and nodes labelled by terms we case on.  Nodes of type $\tfalseC$ have
no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
casing on $t : \NeG{C \lor D}$ have
two subtrees, one having a new hypothesis $C$, and one having a new
hypothesis $C$.

@@

\noindent
We have everything now to model disjunction:
\[
\begin{array}{lcl}
\den{A \lor B} & = & \Cover{\den A \hatplus \den B}
\\[2ex]
\treflect^{A \lor B} & : & \Ne{A \lor B} \todot \den{A \lor B} \\
\reflect[\Gamma]{A \lor B}t & = & \orC t
  {(\returnCP {\injP1 {\reflect[\Gamma.A]A{(\hyp \ttop)}}})}
  {\\ && \qquad\ (\returnCP {\injP2 {\reflect[\Gamma.B]B{(\hyp \ttop)}}})}
\\[2ex]
\treifyOr^{A \lor B} & : & \den{A} \hatplus \den{B} \todot \Nf{A \lor B} \\
\reifyOr[\Gamma]{A \lor B}(\inj1\, a) & = & \orIP1 {\reify[\Gamma]A a} \\
\reifyOr[\Gamma]{A \lor B}(\inj2\, b) & = & \orIP2 {\reify[\Gamma]B b}
\\[1.5ex]
\treify^{A \lor B} & : & \den{A \lor B} \todot \Nf{A \lor B} \\
\treify^{A \lor B} & = & \tpasteNf \comp \Cover{\treifyOr^{A \lor B}} \\
\end{array}
\]
Reification works by first reifying the leaves of the case tree using
$\treifyOr$, and then turning the whole tree into a normal form using $\tpasteNf$.
\begin{exercise}
  Compute $\reify[P \lor Q]{P \lor Q} (\reflect[P \lor Q]{P \lor Q}
  (\hyp \ttop))$.
\end{exercise}
For absurdity, we use the presheaf $\hat0_\Gamma = \{\}$ with initial
morphism $\botelim : \hat0 \todot \C$.
\[
\begin{array}{lcl}
\den{\bot} & = & \Cover{\hat0}
\\[2ex]
\treflect^{\bot} & : & \Ne{\bot} \todot \den{\bot} \\
\reflect[\Gamma]{\bot}t & = & \falseC t
\\[2ex]
\treify^{\bot} & : & \den{\bot} \todot \Nf{\bot} \\
\treify^{\bot} & = & \tpasteNf \comp \Cover{\botelim} \\
\end{array}
\]
To reify a value $c : \den\bot$, observe that because $\hat0$ is
empty, $c$ can only be a case tree without leaves, \ie, all branches
end in $\tfalseC$ nodes.  Thus, the mapping
$\Cover{\botelim} : \Cover{\hat0} \todot \Cover{\Nf\bot}$ is merely a
type cast.  The subsequent $\tpasteNf$ will turn the leafless tree $c$
into a normal form consisting entirely of splits $\torE$ and $\tfalseE$.

@@

Evaluation poses one more challenge. Consider the term:
\[
\ru{x : \HypG {C \lor D} \qquad
    f : \Tm[\Gamma.C]{A \imp B} \qquad
    g : \Tm[\Gamma.D.A]{B} \qquad
    a : \TmG A
  }{\impEP{\orE {(\hyp x)} f {(\impI g)}} a : \TmG B}
\]
There is a redex $\impEP {\impI g} a$ separated by a $\torE$.  The latter
would naturally evaluate to a case tree with functions in the
leaves.  To apply it to the evaluation of $a$, we have to apply each
leaf to that argument.  We facilitate this by implementing
\[
  \tpaste^{A \imp B} : \Cover{\den{A \imp B}} \todot \den{A \imp B}
\]
that turns a case tree of functions into a function, which expects an
argument that will then be passed to all the functions in the case
tree.  In fact, we need pasting at all formulas $A$:
\[
\begin{array}{lcl}
\tpaste^A & : & \Cover{\den A} \todot \den A \\[1ex]
\tpaste^P & = & \tpasteNf \\
\tpaste^\bot & = & \tjoinC \\
\tpaste^{A \lor B} & = & \tjoinC \\
\tpaste^{A \land B} & = & \langle\,
  \tpaste^A \comp \Cover{\pi_1} \, , \;
  \tpaste^B \comp \Cover{\pi_2} \,\rangle \\[1ex]
\multicolumn 3 {l} {\paste[\Gamma]{A \imp B} c \;
  \Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
  (\tpaste^B \comp \mapC \phi \comp \Cover{\den {A \imp B}_\tau}\, c)
}
\end{array}
\]
Here, we need a stronger version of functoriality of $\tCover$ which
only requires an
\[
  \phi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
             \to \B_\Phi
\]
to produce a $\tmapC_\Delta t\; \phi : \Cover\A_\Delta \to \Cover\B_\Delta$.
In our case, $\A = \den{A \imp B}$ and $\B = \den{B}$ and
\[
  \phi\;\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
  .
\]
Pasting allows us to complete the definition of evaluation.
\[
\begin{array}{lcl}
  \teval_\Gamma^A & : & \TmG A \to \den \Gamma \todot \den A \\[1ex]
  \evalG C {\orE t u v} \Delta \; \gamma & = &
    \tpaste^C \;
     (\mapC {(\lambda \Phi\, (\tau  : \Phi \leq \Delta) \to [f,g])}\;c) \\[1ex]
  \multicolumn 1 r {\mbox{ where } c} & = &
      \evalG{A \lor B} t\;\Delta\;\gamma \\
  \multicolumn 1 r {f\;a} & = &
    \eval[\Gamma.A] C u \;\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
  \multicolumn 1 r {g\;b} & = &
    \eval[\Gamma.B] C v \;\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
  \\
\end{array}
\]
\begin{exercise}
  Really complete the definition of $\teval$ (all missing cases).
\end{exercise}
That's it!

@@

\paradot{Acknowledgments}  The presentation of this NbE proof and its
Agda formalization benefitted greately from discussions with Christian
Sattler.  Thanks to Jesper Cockx for listening to a prerun of the talk.

\bibliographystyle{abbrvnat}
\bibliography{auto-nbeSum}

>>>>

\nonstopmode
\documentclass[a4paper]{article}
\usepackage{natbib}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
%\usepackage{dsfont}
\usepackage{bibentry}

@@

\nobibliography{auto-nbeSum}

@@

Indices $x$ into the context $\Gamma$ of hypothesis, locating hypothesis
$A$, are written $x : \Hyp[\Gamma]A$ and defined inductively by the
following constructors:
\[
  \ru{}{\ttop : \Hyp[\Gamma.A]A}
\qquad
  \ru{x : \Hyp[\Gamma]A
    }{\pop[B] x : \Hyp[\Gamma.B]A}
\]
Derivations $t$ of the truth of a formula $A$ under assumptions
$\Gamma$ (judgement $\Gamma \der A$)
are written $t : \Tm[\Gamma]A$ and defined inductively as follows:

@@

The following indexed grammar for \emph{normal} derivations $\NfG A$ of
judgement $\Gamma \der A$ ensures the subformula property.  It is
defined mutually with a grammar $\NeG A$ of \emph{neutral} derivations.
Neutral derivations are ``straight-line'' consequences of hypotheses
(without case distinction or absurdity elimination).

@@

\begin{exercise}
  Using the constructions in this section, show that the category of
  presheaves and natural transformations is Cartesian closed.
\end{exercise}

@@

Instead of defining the semantics $\den A$ of formula $A$ by induction
on $A$, we can consider a category whose objects are interpretations
of formulas, and show that this category is Cartesian closed.

An object in this category would be a quadruple consisting of:
\[
\begin{array}{lll@{\qquad}l}
  \A & : & \PSh  & \mbox{a presheaf} \\
  A  & : & \Form & \mbox{a formula} \\
  \treflect^A & : & \Ne A \todot \A
    & \mbox{a method to reflect neutrals into $\A$} \\
  \treify^A & : & \A \todot \Nf A
    & \mbox{a method to reify values of $\A$} \\
\end{array}
\]
\begin{exercise}
  Complete this definition by a suitable notion of morphism.
\end{exercise}
\begin{exercise}
  Show that this category is Cartesian closed.
\end{exercise}

@@

Given a presheaf $\A$, we define a new presheaf $\Cover \A$
inductively by the following constructors:
\begin{gather*}
  \ru{a : \A_\Gamma}{\returnC a : \Cover\A_\Gamma}
\qquad
  \ru{t : \NeG \bot}{\falseC t : \Cover\A_\Gamma}
\\[1ex]
  \ru{t : \NeG{C \lor D} \qquad
      c : \Cover\A_{\Gamma.C} \qquad
      d : \Cover\A_{\Gamma.D}
    }{\orC t c d : \Cover\A_\Gamma}
\end{gather*}
\begin{exercise}
  Show that $\Cover\A$ is indeed a presheaf.
\end{exercise}
We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
and nodes labeled by terms we case on.  Nodes of type $\tfalseC$ have
no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
casing on $t : \NeG{C \lor D}$ have
two subtrees, one having a new hypothesis $C$, and one having a new
hypothesis $C$.

@@

\noindent
We have everything now to model disjunction:
\[
\begin{array}{lcl}
\den{A \lor B} & = & \Cover{\den A \hatplus \den B}
\\[2ex]
\treflect^{A \lor B} & : & \Ne{A \lor B} \todot \den{A \lor B} \\
\reflect[\Gamma]{A \lor B}t & = & \orC t
  {(\returnCP {\injP1 {\reflect[\Gamma.A]A{(\hyp \ttop)}}})}
  {\\ && \qquad\ (\returnCP {\injP2 {\reflect[\Gamma.B]B{(\hyp \ttop)}}})}
\\[2ex]
\treifyOr^{A \lor B} & : & \den{A} \hatplus \den{B} \todot \Nf{A \lor B} \\
\reifyOr[\Gamma]{A \lor B}(\inj1\, a) & = & \orIP1 {\reify[\Gamma]A a} \\
\reifyOr[\Gamma]{A \lor B}(\inj2\, b) & = & \orIP2 {\reify[\Gamma]B b}
\\[1.5ex]
\treify^{A \lor B} & : & \den{A \lor B} \todot \Nf{A \lor B} \\
\treify^{A \lor B} & = & \tpasteNf \comp \Cover{\treifyOr^{A \lor B}} \\
\end{array}
\]
Reification works by first reifying the leaves of the case tree using
$\treifyOr$, and then turning the whole tree into a normal form using $\tpasteNf$.
\begin{exercise}
  Compute $\reify[P \lor Q]{P \lor Q} (\reflect[P \lor Q]{P \lor Q}
  (\hyp \ttop))$.
\end{exercise}
For absurdity, we use the presheaf $\hat0_\Gamma = \{\}$ with initial
morphism $\botelim : \hat0 \todot \C$.
\[
\begin{array}{lcl}
\den{\bot} & = & \tCover\,{\hat0}
\\[2ex]
\treflect^{\bot} & : & \Ne{\bot} \todot \den{\bot} \\
\reflect[\Gamma]{\bot}t & = & \falseC t
\\[2ex]
\treify^{\bot} & : & \den{\bot} \todot \Nf{\bot} \\
\treify^{\bot} & = & \tpasteNf \comp \Cover{\botelim} \\
\end{array}
\]
To reify a value $c : \den\bot$, observe that because $\hat0$ is
empty, $c$ can only be a case tree without leaves, \ie, all branches
end in $\tfalseC$ nodes.  Thus, the mapping
$\Cover{\botelim} : \Cover{\hat0} \todot \Cover{\Nf\bot}$ is merely a
type cast.  The subsequent $\tpasteNf$ will turn the leafless tree $c$
into a normal form consisting entirely of splits $\torE$ and $\tfalseE$.

@@

Evaluation poses one more challenge. Consider the term:
\[
\ru{x : \HypG {C \lor D} \qquad
    f : \Tm[\Gamma.C]{A \imp B} \qquad
    g : \Tm[\Gamma.D.A]{B} \qquad
    a : \TmG A
  }{\impEP{\orE {(\hyp x)} f {(\impI g)}} a : \TmG B}
\]
There is a redex $\impEP {\impI g} a$ separated by a $\torE$.  The latter
would naturally evaluate to a case tree with functions in the
leaves.  To apply it to the evaluation of $a$, we have to apply each
leaf to that argument.  We facilitate this by implementing
\[
  \tpaste^{A \imp B} : \Cover{\den{A \imp B}} \todot \den{A \imp B}
\]
that turns a case tree of functions into a function, which expects an
argument that will then be passed to all the functions in the case
tree.  In fact, we need pasting at all formulas $A$:
\[
\begin{array}{lcl}
\tpaste^A & : & \tCover\,{\den A} \todot \den A \\[1ex]
\tpaste^P & = & \tpasteNf \\
\tpaste^\bot & = & \tjoinC \\
\tpaste^{A \lor B} & = & \tjoinC \\
\tpaste^{A \land B} & = & \langle\,
  \tpaste^A \comp \Cover{\pi_1} \, , \;
  \tpaste^B \comp \Cover{\pi_2} \,\rangle \\[1ex]
\multicolumn 3 {l} {\paste[\Gamma]{A \imp B} c \;
  \Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
  (\tpaste^B \comp \mapC \phi \comp \Cover{\den {A \imp B}_\tau}\, c)
}
\end{array}
\]
Here, we need a stronger version of functoriality of $\tCover$ which
only requires an
\[
  \phi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
             \to \B_\Phi
\]
to produce a $\tmapC_\Delta t\; \phi : \Cover\A_\Delta \to \Cover\B_\Delta$.
In our case, $\A = \den{A \imp B}$ and $\B = \den{B}$ and
\[
  \phi\;\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
  .
\]
Pasting allows us to complete the definition of evaluation.
\[
\begin{array}{lcl}
  \teval_\Gamma^A & : & \TmG A \to \den \Gamma \todot \den A \\[1ex]
  \evalG C {\orE t u v} \Delta \; \gamma & = &
    \tpaste^C \;
     (\mapC {(\lambda \Phi\, (\tau  : \Phi \leq \Delta) \to [f,g])}\;c) \\[1ex]
  \multicolumn 1 r {\mbox{ where } c} & = &
      \evalG{A \lor B} t\;\Delta\;\gamma \\
  \multicolumn 1 r {f\;a} & = &
    \eval[\Gamma.A] C u \;\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
  \multicolumn 1 r {g\;b} & = &
    \eval[\Gamma.B] C v \;\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
  \\
\end{array}
\]
\begin{exercise}
  Really complete the definition of $\teval$ (all missing cases).
\end{exercise}
That's it!

\newcommand{\citeentry}[1]{\item\nocite{#1}\bibentry{#1}}

\section{Literature}

Further reading on normalization by evaluation in general:
\begin{enumerate}

\citeentry{bergerSchwichtenberg:lics91}

\citeentry{jungTiuryn:tlca93}

\citeentry{altenkirchHofmannStreicher:ctcs95}

\citeentry{abel:habil}

\end{enumerate}

% \begin{itemize}
% \item \citet{bergerSchwichtenberg:lics91}
% \item \citet{jungTiuryn:tlca93}
% \item \citet{altenkirchHofmannStreicher:ctcs95}
% \item \citet{abel:habil}
% \end{itemize}
\noindent
Literature dealing in particular with the case of disjunction /
disjoint sums:

\begin{enumerate}

\citeentry{danvy:popl96}

\citeentry{fioreSimpson:tlca99}

\citeentry{altenkirchDybjerHofmannScott:lics01}

\citeentry{balatDiCosmoFiore:popl04}

\citeentry{altenkirchUustalu:flops04}

\citeentry{barral:PhD}

\citeentry{scherer:popl17}

\end{enumerate}

% \begin{itemize}
% \item \citet{danvy:popl96}
% \item \citet{fioreSimpson:tlca99}
% \item \citet{altenkirchDybjerHofmannScott:lics01}
% \item \citet{balatDiCosmoFiore:popl04}
% \item \citet{altenkirchUustalu:flops04}

% \item \citet{barral:PhD}
% \item \citet{scherer:popl17}
% \end{itemize}

@@

\paradot{Acknowledgments}  The presentation of this NbE proof and its
Agda formalization benefited greatly from discussions with Christian
Sattler.  Thanks to Jesper Cockx for listening to a prerun of the talk.

\bibliographystyle{abbrvnat}
%\bibliography{auto-nbeSum}
